What you see is what you write

### 2018-10-20 wip-live-programming/paper.tex
Uncommitted changes

\usepackage{lineno,hyperref,xcolor}
\usepackage{flushend}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{booktabs} 
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage[shortlabels]{enumerate}
\setlist{leftmargin=6mm}

@@

\title{Design and implementation of a live coding environment for data science}

\author{Tomas Petricek}
%\affiliation{
%  \institution{The Alan Turing Institute}
%  \country{London, United Kingdom}
%}
%\email{tomas@tomasp.net}

@@

\begin{document}
\begin{abstract}
Data science can be done by directly manipulating data using spreadsheets, or by writing 
data manipulation scripts using a programming language. The former is error-prone and
does not scale, while the latter requires expert skills. Live coding has the potential to 
bridge this gap and make writing of transparent, reproducible scripts more accessible.

In this paper, we describe a live programming environment for data science that provides instant 
previews and contextual hints, while allowing the user to edit code in an unrestricted way in
a text editor.

Supporting a text editor is challenging as any edit can significantly change the structure of code and
fully recomputing previews after every change is too expensive. We present a technique that allows the type
checker and the interpreter to run on the fly, while code is written, and reuse results of previous 
runs. This makes it possible to efficiently provide instant feedback and live previews during development.

We formalise how programs are interpreted and how previews are computed, prove correctness
of the previews and formally specify when can previews be reused. We believe this work provides
solid and easy to reuse foundations for the emerging trend of live programming environments.
\end{abstract}
\maketitle

@@

One of the aspects that make spreadsheet tools such as Excel more accessible than programming 
environments is their liveness. When you change a value in a cell in Excel \cite{spreadsheet}, the  
whole spreadsheet is updated instantly and you see the new results immediately.

Increasing number of programming environments aim to provide the same live experience for more
standard programming languages, but doing this is not easy. Fully recomputing the whole program after
every single change is inefficient and calculating how a change in source code changes the result
is extremely hard when the editor allows arbitrary manipulation of program text. For example, 
consider the following simple program that gets the release years of 10 most expensive movies 
in a data set \ident{movies}:
%
\begin{equation*}
\begin{array}{l}  
\kvd{let}~\ident{top} =\ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\num{10})\\
\quad .\ident{map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{yyyy}))}\\
\end{array}
\end{equation*}
%
A live coding environment shows a preview of the list of dates. Next, assume that the programmer
modifies the code by making the constant $\num{10}$ a variable and changing the date format to see 
the full date:
%
\begin{equation*}
\begin{array}{l}  
\kvd{let}~\ident{count} = \num{10}\\
\kvd{let}~\ident{top} = \ident{movies}\\
\quad .\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())\ident{.take}(\ident{count})\\
\quad \ident{.map}(\lambda x \rightarrow x.\ident{getReleased()}.\ident{format}(\str{dd-mm-yyyy}))}\\
\end{array}
\end{equation*}
%
Ideally, the live coding environment should understand the change, reuse a cached result of the
first two transformations (sorting and taking 10 elements) and only evaluate the last \ident{map}
to differently format the release dates of already computed top 10 movies.

This is not difficult if we represent the program in a structured way \cite{structure-based,hazelnut} 
and allow the user to edit code via known operations such as ``extract variable'' (which has
no effect on the result) or ``change constant value'' (which forces recomputation of subsequent
transformations). However, many programmers prefer to edit programs as free form text.

@@

>>>>

\usepackage{lineno,hyperref,xcolor}
\usepackage{ragged2e}
\usepackage{flushend}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage{xypic}
\usepackage{semantic}
\usepackage{booktabs} 
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage[shortlabels]{enumerate}
\setlist{leftmargin=6mm}

@@

\title{What you see is what you write}
\subtitle{On the fly evaluation of live previews for data manipulation scripts}

\author{Tomas Petricek}
\affiliation{
  \institution{The Alan Turing Institute}
  \country{London, United Kingdom}
}
\email{tomas@tomasp.net}

@@

\begin{document}
\begin{abstract}
Liveness is one of the aspects that make spreadsheets easy to use. When you edit a formula in 
Excel, the whole spreadsheet is automatically updated and you see the result immediately. 
Providing the same behaviour for data manipulation scripts would make simple data scripting 
more accessible, but it is difficult to do efficiently while allowing users to edit code in 
an unrestricted way in a text editor.

In this paper, we describe an efficient method for evaluating live previews for simple data 
manipulation scripts. This is challenging as any edit in a text editor can significantly 
change the structure of code and fully recomputing previews after every change is too expensive. 
Our method allows the interpreter to run while code is written and reuse results of previous 
runs. This makes it possible to efficiently provide live previews on the fly during development.

We formalise how live previews are computed, prove their correctness and formally specify when 
a live preview can be reused. We also demonstrate the efficiency of our method using an empirical 
study. Although our focus is on a simple data scripting language, the work provides easy to use 
foundations that can be shared by a wide range of live programming environments.
\end{abstract}
\maketitle

@@

Data science tools such as Jupyter notebooks \cite{jupyter} and programming environments
with REPLs \cite{rmarkdown,drscheme} embed program outputs directly into the environment, 
but the outputs are only updated after the user finishes writing code and explicitly runs it.
Doing better is not easy. Fully recomputing the whole program after every single key stroke is 
inefficient and calculating how a change in source code changes the result is extremely hard 
when the editor allows arbitrary manipulation of program text.

For example, consider two steps of a data manipulation process that are documented in 
Figure~\ref{fig:intro-live}. The user first writes code to load and aggregate broadband speed data 
(described in Appendix~\ref{sec:app-broadband}). Using a live preview, she spots a problem with 
the data and decides to fix it by ignoring invalid rows. She then changes the code by adding
a call to filter the data and also extracts a constant into a named variable.

\begin{figure}[b]
\caption{\small{Exploring broadband speed data. We group reports by whether they are from urban
  or rural area and calculate average download speeds. Resulting table is displayed as we type.}}
\label{fig:intro-live}

\vspace{0.5em}
\noindent
\includegraphics[scale=0.38]{bb1.png}

\vspace{0.5em}
\justify
\small{We notice that Urban/rural may be empty and filter out empty values and we
  extract the URL into a variable. To update live preview, the system needs to evaluate
  only the last filtering operation.}

\vspace{0.75em}
\noindent
\includegraphics[scale=0.38]{bb2.png}
\end{figure}

A system that provides live previews should understand the code change, reuse cached result of data 
loading and aggregation and only evaluate the filtering operation. Regardless of the size of the 
original input, this will be fast as it filters an aggregated series with just three rows.

This is not difficult if we represent the program in a structured way \cite{structure-based,hazelnut} 
and allow the user to edit code via known operations such as ``extract variable'' (which has
no effect on the result) or ``add operation to a chain'' (which requires evaluation of the new
transformation). However, many programmers prefer to edit programs as free form text.

@@

\section{Appendix}
\label{sec:app-broadband}
